Practical consistency management for geographically distributed MMOG servers

### 2011-05-31 trunk/textos/optpaxos/draftv2.tex

\newcommand{\blankline}{\vspace{4 mm}}

\newcommand{\cmsend}[1]{GFTO-Mcast({#1})}

@@

To ensure reliability despite server crashes, each server is replicated using state-machine replication, implemented with Paxos. Each player sends his commands to the server to which he is connected, which then proposes that command in a consensus instance. Each command is assigned a timestamp and executed against objects in timestamp order. We implement this by using a logical clock in each server group. Guaranteeing that the same set of commands is executed by the respective affected objects in the same order provides the level of consistency we are seeking. Therefore, the challenge is how to assign timestamps to commands such that consistency is not violated and commands are not discarded due to stale timestamp values.

@@

Here, we consider that there are several processes. Each process $p$ belongs to a group $G = group(p)$, that is, $p \in G = group(p)$. Each group $G$ has a special leader process $c$, which is its coordinator.

@@

The basic idea of the optimistic ordering is the following: assuming that the processes have a synchronized clock\footnote{We don't require here perfectly synchronized clocks, as the optimistic protocol tolerates mistakes by its very definition. We only need clocks which are synchronized enough, so that our delivery order prediction succeeds and matches the conservative delivery order.}, whenever a process $p$ (either coordinator or not) receives a message $m$ from a client, it immediately applies a timestamp $ts$ to it, which consists simply of the current value of $p$'s wallclock, $now$. Therefore, $m.ts=now$. A wait window of length $w(p)$ is considered, where $w(p)$ is defined as the highest value of the estimated communication delay plus the wallclock deviation between the process $p$ and any of the other processes in its group $G$ or in any of the groups in $sendersTo(G)$.

More formally, let $\delta(p',p)$ be the maximum time for a message from $p'$ to arrive at $p$. Also, let $\epsilon(p',p)$ be the deviation between the wallclocks of $p$ and $p'$. The value of $w(p)$ is estimated as the maximum value of $(\delta(p',p) + \epsilon(p',p))$ for every $p'$ in $G = group(p)$ or in some group of $sendersTo(G)$.

@@

To avoid out-of-order deliveries, when a message $m$ arrives too late, that is, if when $m$ arrives at $p$, $m.ts < now - w(p)$, the message is simply discarded by $p$, since another message $m' : m'.ts > m.ts$ may have already been delivered\footnote{We could make this in a way such that $m$ is only discarded by $p$ if, in fact, there was a delivered message $m' : m'.ts > m.ts$. If there was no such message, $m$ could still be delivered without violating the order we defined.}.

@@

  In this case, when $m$ is optimistically delivered by $p$, $p$ checks whether the latest consensus instance $I_{prev}$ in which it participated, or is trying to start, has already been decided -- if not, $p$ enqueues $m$ in a queue $propPending$ as the next message being proposed by it, so other tasks can keep being executed. Then, once $I_{prev}$ has been decided, $p$ checks wheter $m$ -- or a message $m' : m'.ts > m.ts$ -- has already been conservatively delivered in some previous consensus instance. If such message was indeed delivered previously, $p$ discards $m$; otherwise $p$ proposes $m$ as the next message to be conservatively delivered by the processes of $G$. When $m$ is conservatively delivered, it is not immediately delivered to the application. Instead, it is inserted in a $consPending$ list for later being conservatively delivered, which will happen once every group $G'$ in $sendersTo(G)$ have already sent a message $barrier(G',G) = t$, such that $t > m.ts$. This is done because there could be a message $m'$ yet to come from another group $G'$, such that $m'.ts < m.ts$.

@@

  In this case, when $now > m.ts + w(p)$, $p$ tries to initiate a consensus instance within $G$ to decide $m$. If $p$ cannot start the proposal now, $m$ is enqueued in $propPending$ for being proposed later. Before proposing $m$, $p$ checks whether $m$ -- or a message $m' : m'.ts > m.ts$ -- has already been decided in some previous consensus instance. If it has, $m$ is discarded; otherwise, $p$ starts a consensus instance proposing $m$. Once any message $m$ is decided, if $G \in m.dst$, $m$ is inserted in the $consPending$ list. Besides, when $m$ is decided, $p$ sends a message $\{m, \text{`cons'}\}$ to every \mbox{$p' : (\exists G' \in m.dst \setminus \{G\}) \wedge (p' \in G')$}. When $\{m, \text{`cons'}\}$ is received by each $p' \in G'$, $p'$ checks whether it has ever inserted $m$ in its own $consPending$ list. If not, $p'$ inserts $m$ into $consPending$ and adjusts $barrier(G, G')$ to $m.ts$. To ensure that, once a message $\{m, \text{`cons'}\}$ is received, there is no other message \mbox{$\{m', \text{`cons'}\}:m'.ts<m.ts$} coming from $m.src$, every $\{m, \text{`cons'}\}$ message is sent through a lossless FIFO channel\footnote{An ordinary TCP connection would be enough to provide such FIFO lossless channel.}.

@@

Moreover, each process $p$ of group $G$ keeps an \textit{optPending} list which contains the messages waiting to be \textit{OPT-Delivered} (optimistically delivered). The \mbox{\textit{consPending}} list contains the messages ready to be \textit{CONS-Delivered} (conservatively delivered), but which may be waiting for the clearance from the groups in $sendersTo(G)$. Also, each message $m$ has at least four fields: $dst$, which is the set of $m$'s destination groups; $src$, which is the group where $m$ was generated; $ts$, the value of the wallclock of the process when $m$ was generated; and $gs$, which is $m$'s sequence number relatively to its group of origin. Note that, by abuse of notation, we have `process $p \in m.dst$' instead of `process $p$ : $\exists$ group $G \in m.dst \wedge p \in G$'.

Something that must be noticed is that some messages might not have their source groups as a destinations. Anyway, each message $m$ of this type still has to be agreed upon in its groups of origin, so that $m.gs$ can be decided, and also $m$ will be retrievable, even in the presence of failures.% Because of that, to ensure that messages of this kind are also received by $group(p)$'s coordinator $c$, which then initiates the Paxos instance that allows for the conservative delivery of the message in all of its destinations, every message $m$ from $p$ must also be sent to $c$ (l. \ref{algline:sendcoord}). This way, when the clock value at the coordinator $time(c) > w(c) + m.ts$, $c$ will conservatively send $m$ to all the processes in $m.dst$. Finally, to avoid delivering a message $m$ in a group it is not addressed to, before OPT-Delivering it, each process must check whether it is one of $m$'s destinations or not (l. \ref{algline:notinterested}). In l. \ref{algline:sendcoord}, $m$ is sent to the whole group $G$, so that in the case of failure of $G$'s coordinator, it is more likely that the new coordinator will know about the existence of $m$.

@@

\blankline
\INDENT{\textit{When} Decide($k$,$m$)}
%  \STATE $m.gs \leftarrow k$
  \STATE $decided \leftarrow decided \cup \{m\}$
  \STATE $propPending \leftarrow propPending \setminus (\{m\} \cup \{m' : m'.ts < m.ts\})$
  \STATE $nextProp \leftarrow k + 1$
  \STATE $k \leftarrow k + 1$  
  \IF {$G \in m.dst$} \label{algline:checkcons}
    \STATE $consPending \leftarrow consPending \cup \{m\}$
  \ENDIF
  \INDENT{\textbf{for all} $p' \in m.dst \setminus \{G\}$ \textbf{do}}
    \STATE send($p'$, $\{m, \text{`cons'}\}$)    
  \ENDINDENT
\ENDINDENT

\blankline
\INDENT{\textit{When} receive($\{m',\{\text{`cons'}\}$)}
  \STATE $barrier(m'.dst,G) \leftarrow max(m'.ts, barrier(m'.dst,G))$
  \IF {$m' \notin\ consPending \wedge\ m' \notin\ delivered \wedge\ m' \neq\ null$}
    \STATE $consPending \leftarrow consPending \cup \{m\}$
  \ENDIF
\ENDINDENT

@@

\blankline
\INDENT{\textit{When $\exists G' \in receiversFrom(G):$\\$getTime() - lastBarrierSent(G') > barrierThreshold$}}
  \STATE $null \leftarrow$ empty message
  \STATE $null.ts \leftarrow getTime()$
  \STATE $null.src \leftarrow G$
  \STATE $null.dst \leftarrow G'$
  \STATE \cmsend{$null$}
  \COMMENT{saving that nothing was sent until null.ts}
\ENDINDENT

@@

The problem with addressing liveness this way is that, in the worst case, $G'$ has decided a message $m$ and has just received the last barrier $b$ from $G$, such that $b$ is less than $m.ts$ by an infinitesimal difference. This would mean that, if $G$ has no messages to send to $G'$, $G'$ will have to wait at least for $barrierThreshold + T(consensus)$, where $T(consensus)$ is the time needed for a value to be decided after it has been proposed in some consensus instance -- assuming that such value is going to be decided at all in that instance. Because of that, the conservative delivery of $m$ will be delayed by $w(p) + barrierThreshold + 2T(consensus)$, where $w(p)$ is the wait window of the process $p \in G$, and $T(consensus)$ is counted twice because there are two consensus proposals -- first in $G'$ regarding $m$, and then in $G$ regarding the empty message.

There is a way to provide liveness withouth such a possibly big delay, although creating more messages and making deeper changes in the delivery algorithm. Let $blockers(m)$ be defined as the set of groups whose barrier is needed in order to some group to deliver $m$. More formally $blockers(m) = \{G_B : \exists G_{dst} \in m.dst \wedge G_B \in sendersTo(G_{dst})\}$. The idea is that, once each group $G_B$ in $blockers(m)$ have sent a barrier $b > m.ts$ to all the groups in $m.dst \cap receiversFrom(G_B)$, all possible destinations of $m$ can deliver it. This way, instead of relying on periodic messages, whenever a process $p$ in a group $G$ has a message $m$ to send, $p$ sends $m$ to every $p' \in m.dst \cup G \cup blockers(m)$. It is sent to the groups in $blockers(m)$, so that they know that there is a message which will be blocked until they send a proper barrier to unblock it.

When the process $p'$ of some group $G'$ receives $m$, such that $m.src \neq G'$, $p'$ knows that there might be other groups depeding on the barrier of $G'$ to deliver $m$. For that reason, it will immediately create a $null$ message with a timestamp equals to $m.ts$ and with $null.dst = m.dst$ $\cap$ $receiversFrom(G')$. Then, $p$ will execute \mbox{\cmsend{$null$}}. As soon as $null.ts > now - w(p')$, the $null$ message will be proposed in $G'$ and, once $null$ is decided, each process $p'$ of $G'$ will send a $\{null, \text{`cons'}\}$ message to every process in every group \mbox{$G \in m.dst \cap receiversFrom(G')$} -- this way, any group which was waiting for a barrier from $G'$ to deliver $m$ will be able to do so as soon as it receives such a message. The new delivery algorithm would be as described in Algorithm {\ref{algorithm:ondemand}}.

\begin{algorithm}
\begin{distribalgo}[1]
%\blankline\
%\STATE\underline{This algorithm is executed by every process $p$, which belongs to group $G$:}
%\STATE
\blankline
\INDENT {Initialization}
  \INDENT{\textbf{for all} $G' \in sendersTo(G)$ \textbf{do}}
    \STATE $barrier(G',G) = -\infty$
  \ENDINDENT
\ENDINDENT

\blankline
\INDENT{\textit{When $p$ generates a message $m$}}
  \STATE $m.ts \leftarrow getTime()$  
  \COMMENT{current wallclock value as the timestamp of $m$}
  \INDENT{\textbf{for all} $p' \in m.dst \cup \{G\} \cup blockers(m)$ \textbf{do}} \label{algline:askclearance}
    \STATE $send(p', m)$
  \ENDINDENT
\ENDINDENT

\blankline
\INDENT{\textit{When a process $p$ receives a message $m'$}}
  \IF {$G \neq m'.src$} \label{algline:askedforclearance}
    \STATE $null \leftarrow$ empty message
    \STATE $null.src \leftarrow G$
    \STATE $null.ts \leftarrow m'.ts$
    \STATE $null.dst \leftarrow m'.dst \cap receiversFrom(G)$
    \STATE $optPending \leftarrow optPending \cup \{null\}$
  \ENDIF
  \IF {$m'.ts < getTime() - w(p) \vee G \notin m'.dst$}
    \STATE discard $m'$
    \COMMENT{late commands probably lead to out-of-order delivery}
  \ELSE
    \STATE $optPending \leftarrow optPending$ $\cup$ $\{m'\}$
  \ENDIF
\ENDINDENT

\blankline
\INDENT{\textit{When $\exists m \in optPending : getTime() > m.ts + w(p)$ $\wedge$\\$\nexists m' \in optPending: m'.ts < m.ts$}}
  \STATE $optPending \leftarrow optPending \setminus \{m\}$
  \IF {$p$ is a coordinator $\wedge$ $m.src = G$}
    \STATE $Propose(m, m.dst)$ %\label{algline:propg}
  \ENDIF
  \IF {$m \neq null$}  %\label{algline:notinterested}
    \STATE OPT-Deliver($m$)  
%   \INDENT{\textbf{for all} $cli \in clients(s)$ \textbf{do}}
%     \STATE $send(cli, \{c, \text{`optimistic'}\})$    
%     \COMMENT{s sends $c$ to its clients, warning it is optimistic}
%   \ENDINDENT
  \ENDIF  
\ENDINDENT

\blankline
\INDENT{\textit{When} $Learn(m)$}
  \IF {$m \neq null$} %\label{algline:checkcons}
    \STATE $consPending \leftarrow consPending \cup \{m\}$
  \ENDIF
  \IF {$m.src \neq G$}
    \STATE $barrier(m.src, G) = m.ts$
  \ENDIF
%   \INDENT{\textbf{for all} $cli \in clients(s)$ \textbf{do}}
%     \STATE $send(cli, \{c, \text{`final'}\})$    
%     \COMMENT{s sends $c$ to its clients, telling it is the final order}
%   \ENDINDENT
%   \IF {$p$ is a coordinator}
%     \STATE $consPending \leftarrow consPending \setminus \{c\}$
%   \ENDIF  
\ENDINDENT

\blankline
\INDENT{\textit{When $\exists m \in consPending : \forall G' \in sendersTo(G): m.ts < barrier(G',G)$ $\wedge$ \\ $\nexists m' \in consPending : m'.ts < m.ts$ }}
  \STATE $consPending \leftarrow consPending \setminus \{m\}$
  \STATE CONS-Deliver($m$)

@@

\caption{\cmsend{m} -- Algorithm executed by every process $p$ from group $G$}
\label{algorithm:ondemand}
\end{distribalgo}
\end{algorithm}

Now, assuming that $w(c)$ is the same for every group coordinator $c$, we have a worst-case delivery latency for $m$ of $w(c) + 2\Delta$, as the $null$ message is created and proposed in parallel with $m$. However, each sender process is sending $m$ to every destination process -- which will not be changed --, but also to every process in its group and to every process belonging to some group of $blockers(m)$. This can create an amount of messages too high to be practical. A solution for that would be the following: instead of sending $m$ to every single process in $G$ and in each group of $blockers(m)$, the sender $p$ sends $m$ only to its destinations and to the coordinators of $G$ and of the groups in $blockers(m)$. That way, the line \ref{algline:askclearance} of Algorithm \ref{algorithm:ondemand} would be:
\\
\\
\textbf{for all} $p' : p' \in m.dst \vee
\\p' \in \{c : c\text{ coordinates }G \vee (\exists G_B \in blockers(m) \wedge c\text{ coordinates }G_B)\}$
\\\textbf{do}

@@

%\bibliographystyle{latex8}
%\bibliography{main}

\end{document}

>>>>

\newcommand{\blankline}{\vspace{4 mm}}
\newcommand{\cmsend}[1]{GFTO-Mcast({#1})}
\newcommand{\tconsm}{T_{cons}}
\newcommand{\tcons}{$\tconsm$}

@@

To ensure reliability despite server crashes, each server is replicated using state-machine replication, implemented with consensus (e.g. Paxos \cite{lamport1998ptp}). Each player sends his commands to the server to which he is connected, which then proposes that command in a consensus instance. Each command is assigned a timestamp and executed against objects in timestamp order. We implement this by using a logical clock in each server group. Guaranteeing that the same set of commands is executed by the respective affected objects in the same order provides the level of consistency we are seeking. Therefore, the challenge is how to assign timestamps to commands such that consistency is not violated and commands are not discarded due to stale timestamp values.

@@

Here, we consider that there are several processes. Each process $p$ belongs to a group $G = group(p)$, that is, $p \in G = group(p)$.% Each group $G$ has a special leader process $c$, which is its coordinator.

@@

The basic idea of the optimistic ordering is the following: assuming that the processes have a synchronized clock\footnote{We don't require here perfectly synchronized clocks, as the optimistic protocol tolerates mistakes by its very definition. We only need clocks which are synchronized enough, so that our delivery order prediction succeeds and matches the conservative delivery order.}, whenever a process $p$ receives a message $m$ from a client, it immediately applies a timestamp $ts$ to it, which consists simply of the current value of $p$'s wallclock, $now$. Therefore, $m.ts=now$. A wait window of length $w(p)$ is considered, where $w(p)$ is defined as the highest value of the estimated communication delay plus the wallclock deviation between the process $p$ and any of the other processes in its group $G$ or in any of the groups in $sendersTo(G)$.

More formally, let $\delta(p',p)$ be the maximum time for a message from $p'$ to arrive at $p$. Also, let $\epsilon(p',p)$ be the deviation between the wallclocks of $p$ and $p'$. The value of $w(p)$ is estimated as the maximum value of \mbox{$(\delta(p',p) + \epsilon(p',p))$} for every $p'$ in $G = group(p)$ or in some group of $sendersTo(G)$.

@@

To avoid out-of-order deliveries, if when $m$ arrives at $p$, \mbox{$m.ts < now - w(p)$}, wich means that $m$ arrived too late, the message $m$ is simply discarded by $p$, since another message $m' : m'.ts > m.ts$ may have already been delivered\footnote{We could make this in a way such that $m$ is only discarded by $p$ if, in fact, there was a delivered message $m' : m'.ts > m.ts$. If there was no such message, $m$ could still be delivered without violating the order we defined.}.

@@

  In this case, when $m$ is optimistically delivered by $p$, $p$ checks whether the latest consensus instance $I_{prev}$ in which it participated, or is trying to start, has already been decided -- if not, $p$ enqueues $m$ in a queue $propPending$ as the next message being proposed by it, so other tasks can keep being executed. Then, once $I_{prev}$ has been decided, $p$ checks wheter $m$ -- or a message $m' : m'.ts > m.ts$ -- has already been decided in some previous consensus instance. If such message has indeed been decided previously, $p$ discards $m$; otherwise $p$ proposes $m$ as the next message from $G$. When $m$ is decided, it is not immediately delivered to the application. Instead, it is inserted in a $consPending$ list for later being conservatively delivered, which will happen once every group $G'$ in $sendersTo(G)$ have already sent a message $barrier(G',G) = t$, such that $t > m.ts$. This is done because there could be a message $m'$ yet to come from another group $G'$, such that $m'.ts < m.ts$.

@@

  In this case, when $now > m.ts + w(p)$, $p$ tries to initiate a consensus instance within $G$ to decide $m$. If $p$ cannot start the proposal now, $m$ is enqueued in $propPending$ for being proposed later. Before proposing $m$, $p$ checks whether $m$ -- or a message $m' : m'.ts > m.ts$ -- has already been decided in some previous consensus instance. If it has, $m$ is discarded; otherwise, $p$ starts a consensus instance proposing $m$. Once any message $m$ is decided, if $G \in m.dst$, $m$ is inserted in the $consPending$ list. Besides, when $m$ is decided, $p$ sends a message $\{m, \text{`cons'}\}$ to every \mbox{$p' : (\exists G' \in m.dst \setminus \{G\}) \wedge (p' \in G')$}. When $\{m, \text{`cons'}\}$ is received by each $p' \in G'$, $p'$ checks whether it has ever inserted $m$ in its own $consPending$ list. If not, $p'$ inserts $m$ into $consPending$ and adjusts $barrier(G, G')$ to $m.ts$. To ensure that, once a message $\{m, \text{`cons'}\}$ is received, every message \mbox{$\{m', \text{`cons'}\}:m'.ts<m.ts$} has already been received from $m.src$, every $\{m, \text{`cons'}\}$ message is sent through a lossless FIFO channel\footnote{An ordinary TCP connection would be enough to provide such FIFO lossless channel.}.

@@

Moreover, each process $p$ of group $G$ keeps an \textit{optPending} list which contains the messages waiting to be \mbox{\textit{OPT-Delivered}} (optimistically delivered); a \mbox{\textit{propPending}} list containing the messages waiting to be proposed by $p$; and a \mbox{\textit{consPending}} list containing the messages ready to be \textit{CONS-Delivered} (conservatively delivered), but which may be waiting for barriers from the groups in $sendersTo(G)$. Also, each message $m$ has at least four fields: $dst$, which is the set of $m$'s destination groups; $src$, which is the group where $m$ was generated; $ts$, the value of the wallclock of the process when $m$ was generated; and $gs$, which is $m$'s sequence number relatively to its group of origin. Note that, by abuse of notation, we have `process \mbox{$p \in m.dst$'} instead of `process \mbox{$p$ : $\exists$ group $G \in m.dst \wedge p \in G$'}.

Something that must be noticed is that some messages might not have their source groups as a destination. Anyway, each message $m$ of this type still has to be agreed upon in its groups of origin, so that $m.gs$ can be decided, and also for $m$ to be retrievable, even in the presence of failures.% Because of that, to ensure that messages of this kind are also received by $group(p)$'s coordinator $c$, which then initiates the Paxos instance that allows for the conservative delivery of the message in all of its destinations, every message $m$ from $p$ must also be sent to $c$ (l. \ref{algline:sendcoord}). This way, when the clock value at the coordinator $time(c) > w(c) + m.ts$, $c$ will conservatively send $m$ to all the processes in $m.dst$. Finally, to avoid delivering a message $m$ in a group it is not addressed to, before OPT-Delivering it, each process must check whether it is one of $m$'s destinations or not (l. \ref{algline:notinterested}). In l. \ref{algline:sendcoord}, $m$ is sent to the whole group $G$, so that in the case of failure of $G$'s coordinator, it is more likely that the new coordinator will know about the existence of $m$.

@@

\blankline
\INDENT{\textit{When} Decide($k$,$m$)}
%  \STATE $m.gs \leftarrow k$
  \STATE $decided \leftarrow decided \cup \{m\}$
  \STATE $propPending \leftarrow propPending \setminus (\{m\} \cup \{m' : m'.ts < m.ts \wedge m \neq null\})$ \label{algline:nullstays} 
  \IF {$G \in m.dst \wedge m \neq null$} \label{algline:checkcons}
    \STATE $consPending \leftarrow consPending \cup \{m\}$
  \ENDIF
  \INDENT{\textbf{for all} $p' \in m.dst \setminus \{G\}$ \textbf{do}}
    \STATE send($p'$, $\{m, \text{`cons'}\}$)
    \COMMENT{this message is sent through a FIFO lossless channel}
  \ENDINDENT
  \STATE $nextProp \leftarrow k + 1$
  \STATE $k \leftarrow k + 1$
\ENDINDENT

\blankline
\INDENT{\textit{When} receive($\{m',\{\text{`cons'}\}$)}
  \IF {$m' \notin\ consPending \wedge\ m' \notin\ delivered \wedge\ m' \neq\ null$}
    \STATE $consPending \leftarrow consPending \cup \{m\}$
  \ENDIF
  \STATE $barrier(m'.src,G) \leftarrow max(m'.ts, barrier(m'.src,G))$
\ENDINDENT

@@

\blankline
\INDENT{\textit{When $\exists G' \in receiversFrom(G):$\\$getTime() - lastBarrierSent(G') > barrierThreshold$}}
  \STATE $null \leftarrow$ empty message
  \STATE $null.ts \leftarrow lastBarrierSent(G') + barrierThreshold$ \label{algline:samenullid}
  \STATE $null.src \leftarrow G$
  \STATE $null.dst \leftarrow G'$
  \STATE \cmsend{$null$}
  \COMMENT{saving that nothing was sent until null.ts}
\ENDINDENT

@@

The problem with addressing liveness this way is that, in the worst case, $G'$ has decided a message $m$ and has just received the last barrier $b$ from $G$, such that $b$ is less than $m.ts$ by an infinitesimal difference. This would mean that, if $G$ has no messages to send to $G'$, $G'$ will have to wait at least for $barrierThreshold + \tconsm$, where \tcons\ is the time needed for a value to be decided after it has been proposed in some consensus instance -- assuming that such value is going to be decided at all in that instance. Because of that, the conservative delivery of $m$ will be delayed by $w(p) + barrierThreshold + 2\tconsm$, where $w(p)$ is the wait window of the process $p \in G$, and \tcons\ is counted twice because there are two consensus proposals -- first in $G'$ regarding $m$, and then in $G$ regarding the empty message.

Another problem with this approach is that it is necessary to guarantee that a $null$ message will eventually be proposed, decided, and a barrier will be sent to some group which might be needing it, so that progression is guaranteed. Therefore, this kind of messages are excluded from the $propPending$ list only when decided in the group (l. \ref{algline:nullstays} of Algorithm \ref{algorithm:quasi}), since if at least one of the processes does not have it, it could never be decided. On the other hand, this brings another problem: if a $null$ message in the $propPending$ list of some process is never removed unless decided, its presence could prevent other messages of being proposed. To circumvent this, we assume that no two different messages have the same timestamp, that is, $m \neq m' \Leftrightarrow m.ts \neq m'.ts$. Now, we only have to ensure that

There is a way to provide liveness withouth such a possibly big delay, although creating more messages and making deeper changes in the delivery algorithm. Let $blockers(m)$ be defined as the set of groups whose barrier is needed in order to some group to deliver $m$. More formally $blockers(m) =$ \mbox{$\{G_B : \exists G_{dst} \in m.dst \wedge G_B \in sendersTo(G_{dst})\}$}. The idea is that, once each group $G_B$ in $blockers(m)$ have sent a barrier $b > m.ts$ to all the groups belonging to \mbox{$m.dst \cap receiversFrom(G_B)$}, all possible destinations of $m$ can deliver it. This way, instead of relying on periodic messages, whenever a process $p$ in a group $G$ has a message $m$ to send, $p$ sends $m$ to every $p' \in m.dst \cup G \cup blockers(m)$. It is sent to the groups in $blockers(m)$, so that they know that there is a message which will be blocked until they send a proper barrier to unblock it.

When the process $p'$ of some group $G'$ receives $m$, such that $m.src \neq G'$, $p'$ knows that there might be other groups depeding on the barrier of $G'$ to deliver $m$. For that reason, it will immediately create a $null$ message with a timestamp equals to $m.ts$ and with $null.dst = m.dst$ $\cap$ $receiversFrom(G')$. Then, $p$ will execute \mbox{\cmsend{$null$}}. As soon as $null.ts > now - w(p')$, the $null$ message will be proposed in $G'$ and, once $null$ is decided, each process $p'$ of $G'$ will send a $\{null, \text{`cons'}\}$ message to every process in every group \mbox{$G \in m.dst \cap receiversFrom(G')$}. This way, any group which was waiting for a barrier from $G'$ to deliver $m$ will be able to do so as soon as it receives such $null$ message. The new delivery algorithm would be as described in \mbox{Algorithm {\ref{algorithm:ondemand}}}.

\blankline
\INDENT{\textit{To send a message $m$} -- \cmsend{m}}
  \STATE $m.ts \leftarrow getTime()$  
  \COMMENT{current wallclock value as the timestamp of $m$}
  \INDENT{\textbf{for all} $p' \in m.dst \cup \{G\} \cup blockers(m)$ \textbf{do}}
    \STATE send($p'$, $m$)    
    \COMMENT{send optimistically $m$ to all involved processes}
  \ENDINDENT
\ENDINDENT

\blankline
\INDENT{\textit{When} receive($m'$)}
  \IF {$G \neq m'.src$}
    \STATE $null \leftarrow$ empty message
    \STATE $null.src \leftarrow G$
    \STATE $null.ts \leftarrow m'.ts$
    \STATE $null.dst \leftarrow m'.dst \cap receiversFrom(G)$
    \STATE $optPending \leftarrow optPending \cup \{null\}$
  \ENDIF
  \IF {$m'.ts < getTime() - w(p) \vee G \notin m'.dst$}
    \STATE discard $m'$
    \COMMENT{late commands probably lead to out-of-order delivery}
  \ELSE
    \STATE $optPending \leftarrow optPending$ $\cup$ $\{m'\}$
  \ENDIF
\ENDINDENT

\blankline
\INDENT{\textit{When} Decide($k$,$m$)}
%  \STATE $m.gs \leftarrow k$
  \STATE $decided \leftarrow decided \cup \{m\}$
  \STATE $propPending \leftarrow propPending \setminus (\{m\} \cup \{m' : m'.ts < m.ts \wedge m' \neq null\})$ \label{algline:keepnull}
  \STATE $nextProp \leftarrow k + 1$
  \STATE $k \leftarrow k + 1$  
  \IF {$G \in m.dst$} \label{algline:checkcons}
    \STATE $consPending \leftarrow consPending \cup \{m\}$
  \ENDIF
  \INDENT{\textbf{for all} $p' \in m.dst \setminus \{G\}$ \textbf{do}}
    \STATE send($p'$, $\{m, \text{`cons'}\}$)    
  \ENDINDENT
\ENDINDENT

\blankline
\INDENT{\textit{When} receive($\{m',\{\text{`cons'}\}$)}  
  \IF {$m' \notin\ consPending \wedge\ m' \notin\ delivered \wedge\ m' \neq\ null$}
    \STATE $consPending \leftarrow consPending \cup \{m\}$
  \ENDIF
  \STATE $barrier(m'.src,G) \leftarrow max(m'.ts, barrier(m'.src,G))$
\ENDINDENT

@@

\caption{\cmsend{m} -- executed by every process $p$ from group $G$}
\label{algorithm:ondemand}
\end{distribalgo}
\end{algorithm}

Now, assuming that $w(p)$ is the same for every process $p$, we have a delivery latency for $m$ of $w(p) + \tconsm$, as the $null$ message is created and proposed in parallel with $m$. However, each sender process is sending $m$ not only to every destination process, but also to every process in its group and to every process belonging to some group of $blockers(m)$. This can create a fairly high amount of messages. Unfortunately, this has to be done to guarantee that such $null$ message will eventually be proposed, decided, and a barrier will be sent to some group which might be needing it. Although this kind of messages are excluded from the $propPending$ list only when decided in the group (l. \ref{algline:keepnull} of Algorithm \ref{algorithm:ondemand}), if at least one of the processes does not have it, it could never be decided.

@@

%\bibliographystyle{latex8}
%\bibliography{main}
\bibliographystyle{ieeetr}
\bibliography{gftommog}
\end{document}
